---
title: "Calculate cumulative proximity for geographic areas"
author: "Sarah Rockhill"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{calc-proximity-to-hazards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  rmarkdown.html_vignette.check_title = FALSE
)
```

This vignette describes how the `get_proximity()` function can be used to calculate cumulative proximity to select features for a geographic area, such as a county, census tract, or neighborhood. The `get_proximity()` function returns a vector of numeric values representing cumulative proximity, or how close a community is to the features of interest. 

Cumulative proximity:
- may be used to describe potential population-level exposure risk to environmental hazards
- is calculated by summing the inverse distances between the geometric center of each area and all features within a given search distance
- may be weighted based on the importance of each feature

## View hazprox datasets

In this example, we will calculate and analyze the cumulative proximity to fires occurring in 2019 for census tracts in the state of Georgia. A simple feature dataset representing all 2020 census tracts in the State of Georgia is LazyLoaded with `hazprox` and can be accessed directly as shown below. Let's inspect our census tract dataset, `ga` by viewing the first five records. The Georgia census tracts are provided as a projected sf MULTIPOLYGON object, meaning that we can perform geospatial analysis of these data directly. 

```{r setup, message = FALSE, warning=FALSE}
library(hazprox)
library(dplyr)
library(sf)
library(ggplot2)

#View first 5 census tract records
ga |> slice(1:5)
```

The `hazprox` package also includes a dataset with information about 16,195 fires occurring in the state of Georgia from 2016 through 2020. The fires dataset is derived from the National Interagency Fire Occurrence 6th Edition (1992-2020) dataset. These data are bundled with the `hazprox` package as a comma separated values file (`ga_fires.csv`) to demonstrate how to transform latitudinal and longitudinal coordinates for analysis. 

```{r}
#Import fires dataset
fname <- system.file("extdata/ga_fires.csv", package = "hazprox")
ga_fires <- read.csv(fname)

#View first 5 records in the Georgia fires dataset
ga_fires |>
  select(Event, CauseType, Acres, Owner, Lat, Lon) |>
  slice(1:5)
```

Each record in the `ga_fires` dataset includes latitude (Lat) and longitude (Lon) coordinates representing the location for each fire origin. We will need to convert these records into spatial points before calculating proximity. You can use the `help()` function to view metadata for datasets bundled with hazprox. The metadata includes important information about each dataset including its source, field descriptions, and any processing steps that were performed during dataset preparation.  

```{r, eval=FALSE}
help("ga_fires")
```

The `ga_fires` documentation indicates that the geographic coordinates are in North American Datum of 1983 (NAD 83). We can use the sf library to convert these data to a points simple feature.  

```{r}
fires_sf <- st_as_sf(ga_fires, coords = c("Lon", "Lat"), crs = 4269)
```

## Prepare data for analysis

For this demonstration, we are only interested in fires occurring during 2019, so we will filter the `fires_sf` dataset based on the `Year` column. 

```{r}
fires_sf <- fires_sf |> filter(Year == 2019)
```

The `fires_sf` dataset contains a column for the total number of acres burned (Acres). We can use this information to weight the fires by severity. First, let's examine the distribution of acres burned. 

```{r}
summary(fires_sf$Acres)
```
The number of acres burned shows a highly right-skewed distribution. Most fires burn fewer than seven acres of land; however, the largest fire burned over 400 acres. For this analysis, we will exclude fires less than one acre and create a weight for the number of acres burned. Due to the skewed distribution, we will log-transform acres. 

```{r}
fires_sf <- fires_sf |>
  filter(Acres >= 1) |>
  mutate(severity  = log(Acres))
```

## Visualize fires 

Let's take a look at where fires are located throughout Georgia by plotting the fire coordinates on a map of Georgia. We will vary the size and color of each point by the number of acres burned to highlight the most severe fires. 

```{r, fig.dim=c(6, 5)}
fires_sf |>
  arrange(severity) |>
  ggplot() +
  geom_sf(data = ga, color = "gray", fill = "white") +
  geom_sf(aes(size = severity, color = severity), alpha = 0.3) +
  scale_color_distiller(palette = "Spectral") +
  labs(title = "Fires (1+ Acres) Occuring in Georgia, 2019") +
  theme_minimal()
```

Fires are well dispersed throughout the state with the exception of the Atlanta metro area. The largest fires appear to be clustered within the southwestern quadrant of the state. 

## Calculate proximity 

Next, let's examine how the cumulative proximity is distributed across census tracts in the state using the `get_proximity()` function. We will weight each fire by the log of the total acres burned (Severity). We can also exclude fires occurring more than 20 km from each block group boundary by setting the tolerance to 20. Note, other units of length can be used to specify the tolerance by providing a units argument (e.g., units = 'mi').

Although we converted our fires data to spatial data with an associated geographic coordinate reference system, these data are not currently projected. Typically we want both our polygon features and our hazards to have the same projected CRS, but let’s see what happens if we neglect to project one of our inputs.  

```{r}
wts <- fires_sf$severity
ga$Proximity <- get_proximity(ga, fires_sf, tolerance = 20, weights = wts)
```

When we use the `get_proximity function`, we get the following warning message:

```to is not in a projected CRS.

Projecting to into (+proj=utm +zone=19 +ellps=GRS80 +units=m +no_defs).```

This message warns us that our fires_sf are not projected. The `get_proximity` function has done us the favor of projecting the `fires_sf` into the same CRS as the `ga` layer. Note, this message does not indicate an error. The function will still work when one of the input datasets is not projected because it assumes that the unprojected dataset should have the same projections as the other input. However, if neither input has a projected CRS or if they have different CRS, `get_proximity` will return an error. 

## Descriptive statistics

Now let's build a histogram of the new Proximity field to examine the distribution. 

```{r, fig.dim = c(4, 3)}
ggplot(ga, aes(Proximity)) +
  geom_histogram(fill = "navy", color = "white")
```

Most tracts have relatively low proximity. The median score is `r round(median(ga$Proximity), 3)` However, a few tracts have very high proximity scores. 

## View map of proximity 

We can use `ggplot2` to visualize the geographic distribution of census tracts across Georgia by proximity score. 

```{r, fig.dim = c(6, 5)}
ga |>
  ggplot(aes(fill = Proximity)) +
  geom_sf(color = NA) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "Census Tract Proximity to Fires Burning 1+ Acre",
       subtitle = "Georgia, 2019") +
  theme_minimal()
```

This concludes the `get_proximity` vignette; however, users may consider additional applications of cumulative proximity statistics to public health issues. For example, the proximity score may be used to identify communities or facilities with higher exposure to fire smoke for health education and outreach or may be used to model correlations with health outcomes such as asthma.  
